Bayesian MCMC diagnostics and summarisation using Tracer 1.6

### 2017-07-25 manuscript/tracer16.tex
add example 1

\documentclass{bioinfo}
\copyrightyear{2012}
\pubyear{2012}

@@

% Title must be 150 characters or less
\title[Tracer 1.6]{Bayesian MCMC diagnostics and summarization using Tracer 1.6}

\author[Rambaut \textit{et~al}]{ Andrew Rambaut\,$^{1}$, Alexei J.~Drummond\,$^{2,3}$, Dong Xie\,$^{2,3}$, Marc A.~Suchard\,$^{4,5,6}$}

\address{
$^{1}$Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK\\
$^{2}$Department of Computer Science, University of Auckland, Auckland, NZ\\
$^{3}$Centre for Computational Evolution, University of Auckland, Auckland, NZ\\
$^{4,5}$Departments of Biomathematics and Human Genetics, David Geffen School of Medicine at UCLA, and \\
$^{6}$Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles, USA \\
}

@@

\section{Motivation:}
Computational evolutionary biology, statistical phylogenetics and coalescent-based population genetics are becoming increasingly central to the analysis and understanding of molecular sequence data. 
\section{Results:}
%We describe here a software package, Tracer version 1.6, which ...

\section{Availability:}
Tracer is open-source under the GNU lesser general public license and available at 
\url{http://beast.bio.ed.ac.uk/Tracer} and \url{https://github.com/beast-dev/tracer}.

@@

Bayesian inference of phylogeny \citep{huelsenbeck2001bayesian, felsenstein2004inferring,yang2006computational} has been a popular approach to reconstruct the evolutionary history and discover the evolutionary relationships among organisms that could come from genes, genomes, individuals or species.
The statistical framework is based on the posterior probability of phylogenetic trees given the data, and currently takes the advantage of Markov chain Monte Carlo (MCMC) to promote the computational speed of Bayesian estimation to reach the approximated optimisation.
The posterior summaries, as well as as scientifically relevant marginalisation, %TODO
are therefore critical important to understand a Bayesian phylogenetic analysis.
%However, the full posterior distribution is high-dimensional and often difficult to visualise. 
We then develop a software package Tracer (version 1.6) for visualising and analysing the trace files generated by Bayesian MCMC runs that is the parameter values sampled from the chain. 
It can be used to analyse runs of BEAST \citep{drummond2007beast,drummond2012bayesian}, BEAST2 \citep{bouckaert2014beast2}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes}, LAMARC \citep{kuhner2006lamarc}, Migrate \citep{beerli2006comparison},  and possibly other MCMC programs.
In addition, Tracer will be able to reconstruct the distribution of effective population sizes over time if the population genetic model was involved in the analysis.

@@

Tracer can analyse either a single MCMC log or combine the shared traces from multiple logs as designed in Figure~\ref{fig:tracer}. 
The list of traces in each log are created from the sampling parameters logged during a MCMC process, which include their statistical summaries from their values at all states logged. 
The effective sample size (ESS) is one of important measurements of a sampled continuous parameter, which tells the number of effectively independent draws from the posterior distribution that the Markov chain is equivalent to.

\begin{figure}[ht]
\includegraphics[width=.38\textwidth]{./figures/tracer.pdf}  
\caption{The design of Tracer}
\label{fig:tracer}
\end{figure}

Tracer can handle 3 trace types, also known as variable types \citep{mendenhall2012introduction}: real, integer, and categorical.
The real trace is used to handle quantitative continuous variables, which is the most common type in logged parameters.
The integer is dealing with quantitative discrete variables, and the categorical trace is for qualitative categorical variables. 
% check about defination 
These three types can be automatically assigned when the log file is imported. But they can also be changed later.

Multiple traces can be selected as shown in Figure~\ref{fig:multitrace}, which will overlay the plots for the different traces allowing comparisons to be made. Similarly multiple trace files can be selected as well to compare different runs. If multiple trace files have the same trace names then a ``Combined'' trace will automatically appear, which can be selected as well as the individual trace files.

@@

\begin{itemize}
\item Estimates - this shows the mean, standard deviation, confidence intervals, ESS, and other statistics about the selected trace(s). A frequency distribution will also be plotted for a single selected trace, such as Figure~\ref{fig:4tabs}.a. Interval bars for quantitative traces (real or integer) or violin plots for categorical traces will be drawn for a comparison, such as Figure~\ref{fig:multitrace}, if multiple traces are selected.

@@

\item Joint-Marginal - this only appears if exactly 2 traces are chosen. It then plots one against the other to look at their joint-marginal distribution, for example, in Figure~\ref{fig:4tabs}.c.

@@

Tracer also provides demographic reconstruction resulting in a graphical plot to estimate the distribution of effective population sizes over time, given a MCMC log from a Bayesian framework or additionally with the tree log depending on the model.
For example, ``Demographic Reconstruction" performs reconstructions of the distribution for a number of models, such as constant size, exponential growth \citep{drummond2002estimating}, and logistic growth. %citation?
Besides ``Bayesian Skyline Reconstruction'' reconstructs a generalised Bayesian skyline plot (e.g. Figure~\ref{fig:flu}.a) describing demographic history under canonical scenarios \citep{drummond2005bayesian},
``GMRF Skyride Reconstruction'' and ``SkyGrid Reconstruction'' are respectively available for a temporal smoothing method using the Gaussian Markov random field (GMRF) model  \citep{minin2008smooth} and a generalisation of the GMRF model \citep{gill2012improving}.
``Lineages Through Time'' plots quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?
All these models are available in BEAST \citep{drummond2007beast}.

Tracer recently offers a fairly general solution of looking at conditional posterior distributions. This, specifically, supports for BSSVS forms of model averaging, in which some parameters are only in the likelihood when their sub-model is ``indicated" by some indicator function that is usually a discrete, integer or boolean variable. In that case the posterior of the parameter should not include the states when it was sampled only in the prior because the sub-model it belonged to was ``turned off". This is relevant for EBSP, Random Local Clocks model, microsatellite model averaging and relaxed clock model averaging methods all from my group in last couple of years. It is also relevant for BSSVS in phylogeography as well depending on how the state is logged.

@@

\subsection*{Divergence Dating}

\begin{figure}[ht]
\subfigure[Estimates]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/frequency.pdf}}
\subfigure[Density]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/multiKDE.pdf}}\\
\subfigure[Joint-Marginal]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/joint-marginal.pdf}}
\subfigure[Trace]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/trace.pdf}}
\caption{Image exported from four analysis tabs in Tracer which is used in the ``Divergence Dating'' tutorial.}
\label{fig:4tabs}
\end{figure}

\subsection*{Virus Practical}

>>>>

\documentclass{bioinfo}
\copyrightyear{2017}
\pubyear{2017}

@@

% Title must be 150 characters or less
\title[Tracer 1.6]{Bayesian MCMC diagnostics and summarisation using Tracer 1.6}

\author[Rambaut \textit{et~al}]{ Andrew Rambaut\,$^{1}$, Alexei J.~Drummond\,$^{2,3}$, Dong Xie\,$^{2,3}$, Marc A.~Suchard\,$^{4,5}$}

\address{
$^{1}$Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK\\
$^{2}$Department of Computer Science, University of Auckland, Auckland, NZ\\
$^{3}$Centre for Computational Evolution, University of Auckland, Auckland, NZ\\
$^{4}$Departments of Biomathematics and Human Genetics, David Geffen School of Medicine, UCLA, and \\
$^{5}$Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles, USA \\
}

@@

\section{Motivation:}
%Computational evolutionary biology, statistical phylogenetics and coalescent-based population genetics are becoming increasingly central to the analysis and understanding of molecular sequence data. 
Bayesian inference of phylogeny using Markov chain Monte Carlo (MCMC) plays a central role for understanding the story of evolutions from molecular sequence data.  
The comprehension of posterior distributions from an implementation of this framework is the significant matter.
\section{Results:}
%We describe here a software package, Tracer version 1.6, which ... 
We here present a software package \emph{Tracer} (version 1.6) for visualising and analysing the trace files generated by Bayesian MCMC runs. 
It also provides kernel density estimation, cheap marginal likelihood estimator, 
multivariate analyses (is this reluctant?), %TODO
demographic trajectory reconstruction, conditional posterior distribution summary, etc. 
\section{Availability:}
Tracer is open-source under the GNU lesser general public license and available at 
\url{http://beast.bio.ed.ac.uk/Tracer} and \url{https://github.com/beast-dev/tracer}.

@@

Bayesian inference of phylogeny \citep{huelsenbeck2001bayesian, felsenstein2004inferring,yang2006computational} has been a popular approach to reconstruct the evolutionary history and discover the evolutionary relationships among organisms that could come from genes, genomes, individuals or species.
The statistical framework is based on the posterior probability of phylogenetic trees given the data, and currently takes the advantage of Markov chain Monte Carlo (MCMC) to promote the computational speed of Bayesian estimation to reach an approximated optimisation.
The posterior summaries, as well as as scientifically relevant marginalisation, %TODO
are therefore critically important for understanding a Bayesian phylogenetic analysis.
%However, the full posterior distribution is high-dimensional and often difficult to visualise. 
We then develop a software package Tracer (version 1.6) for this purpose, which analyses Bayesian MCMC log files containing the list of parameters whose values sampled from the chain. 
It can be used to analyse runs of BEAST \citep{drummond2007beast,drummond2012bayesian}, BEAST2 \citep{bouckaert2014beast2}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes}, LAMARC \citep{kuhner2006lamarc}, Migrate \citep{beerli2006comparison},  and possibly other MCMC programs.
In addition, Tracer will be able to redraw the distribution of effective population sizes over time if a demographic model was involved in the analysis.

@@

\begin{figure}[ht]
\includegraphics[width=.38\textwidth]{./figures/tracer.pdf}  
\caption{The system design of Tracer 1.6}
\label{fig:tracer}
\end{figure}

Tracer is able to analyse either a single MCMC log or combine the shared traces from multiple logs as designed in Figure~\ref{fig:tracer}. 
The list of traces in each log are created from the sampling parameters logged during a MCMC process, which include their statistical summaries from their values at all states logged. 
The effective sample size (ESS) is one of important measurements of a sampled continuous parameter, which tells the number of effectively independent draws from the posterior distribution that the Markov chain is equivalent to.

Tracer handles three trace types, also known as variable types \citep{mendenhall2012introduction}: real, integer, and categorical.
The real trace is used to handle quantitative continuous variables, which is the most common type in logged parameters.
The integer is dealing with quantitative discrete variables, and the categorical trace is for qualitative categorical variables. 
% check about definition 
These three types can be automatically assigned when the log file is imported. But they can also be changed later.

The multiple traces can be selected as shown in Figure~\ref{fig:multitrace}, which will overlay the plots for the different traces allowing comparisons to be made. The multiple trace files can be similarly selected as well to compare different runs. If multiple trace files have the same trace names then a ``Combined'' trace will automatically appear, which can be selected as well as the individual trace files.

@@

%TODO simplify, some duplication to 1st example
\begin{itemize}
\item Estimates - this shows the mean, standard deviation, confidence intervals, ESS, and other statistics about the selected trace(s). A histogram presenting the frequency distribution over values will also be plotted for a single selected trace, such as Figure~\ref{fig:4tabs}.a. Interval bars for quantitative traces (real or integer) or violin plots for categorical traces will be drawn for a comparison, such as Figure~\ref{fig:multitrace}, if multiple traces are selected.

@@

\item Joint-Marginal - this only appears if exactly two traces are chosen. It then plots one against the other to look at their joint-marginal distribution, for example, in Figure~\ref{fig:4tabs}.c.

@@

Tracer also provides demographic reconstruction resulting in a graphical plot to estimate the distribution of effective population sizes over time, given a MCMC log from a Bayesian framework or additionally with the tree log depending on the model.
For example, ``Demographic Reconstruction" performs reconstructions of the distribution for a number of models, such as constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, and logistic growth (?). %citation?
Besides ``Bayesian Skyline Reconstruction'' reconstructs a generalised Bayesian skyline plot (e.g. Figure~\ref{fig:flu}.a) describing demographic history under canonical scenarios \citep{drummond2005bayesian},
``GMRF Skyride Reconstruction'' and ``SkyGrid Reconstruction'' are respectively available for a temporal smoothing method using the Gaussian Markov random field (GMRF) model  \citep{minin2008smooth} and a generalisation of the GMRF model \citep{gill2012improving}.
``Lineages Through Time'' (?) plots quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?
All these models are available in BEAST \citep{drummond2007beast}.

%TODO better to re-write
Tracer recently offers a fairly general solution of looking at conditional posterior distributions. This, specifically, supports for BSSVS forms of model averaging, in which some parameters are only in the likelihood when their sub-model is ``indicated" by some indicator function that is usually a discrete, integer or boolean variable. In that case the posterior of the parameter should not include the states when it was sampled only in the prior because the sub-model it belonged to was ``turned off". This is relevant for EBSP, Random Local Clocks model, microsatellite model averaging and relaxed clock model averaging methods all from my group in last couple of years. It is also relevant for BSSVS in phylogeography as well depending on how the state is logged.

@@

\subsection*{Calibrating the tree to estimate TMRCA}

Figure~\ref{fig:4tabs} demonstrates a snapshot of posterior summaries in Tracer for a BEAST \citep{drummond2007beast} run created from a 898 bp alignment of sequences of 12 species of primates (http://beast.bio.ed.ac.uk/tutorials?). %data citation
%using HKY + $\Gamma$ model, uncorrelated lognormal relaxed molecular clock, and Yule prior on the tree.
In this analysis, we furthermore specify three calibrations on the phylogenetic tree to estimate time to the most recent common ancestor (TMRCA) based on our prior fossil knowledge.

The first TMRCA prior, called ``ingroup'', includes all species except the \textit{Lemur} which will form the outgroup in the tree. %monophyletic
The second group is ``Human-Chimp'' that contains only \textit{Homo sapiens} and \textit{Pan}. 
The last consists of everything except \textit{Lemur}, \textit{Saimiri} and \textit{Tarsius} to create the hominoid-cercopithecoid split, which is named as ``HomiCerco''. 
Since we know the most recent common ancestor of humans and chimps may live on about $5-7$ million years ago, a normal distribution centered at 6 million years with a standard deviation of 0.5 million years is assigned to the ``Human-Chimp'' prior. 
Another normal distribution of $24 \pm 0.5$ (million years) is likewise defined for ``HomiCerco''.

\begin{figure}[ht]
\subfigure[Estimates]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/treeRootHeight.pdf}}
\subfigure[Density]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/multiKDE.pdf}}\\
\subfigure[Joint-Marginal]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/joint-marginal-th-ingroup.pdf}}
\subfigure[Trace]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/trace.pdf}}
\caption{Images exported from four analysis tabs in Tracer for calibrating the tree to estimate TMRCA, where the black is ``treeModel.rootHeight'' parameter, blue is TMRCA parameter of ``Human-Chimp'', red is ``ingroup'', and yellow is ``HomiCerco''.}
\label{fig:4tabs}
\end{figure}

Selecting the ``treeModel.rootHeight'' parameter gives the marginal posterior distribution of the age of the root of entire tree. In the example in Figure~\ref{fig:4tabs}.a, we can conclude the root of the tree is most likely between about 39.01 and 63.67 million years old with the mean of 50.48 million years.

According to the kernel density estimation of the two TMRCA parameters (``Human-Chimp'' in blue and ``HomiCerco'' in yellow) in Figure~\ref{fig:4tabs}.b, these two parameters that we used to calibrate the tree explicitly have posterior distributions very similar to the prior distributions that we specified.

The marginal posterior densities of ``treeModel.rootHeight'' and ``ingroup'' are nearly overlapped each other. If we switch to Figure~\ref{fig:4tabs}.c, their joint-marginal distribution plot is almost lined up together, which indicates ??. But this isn't an issue here, because the ``ingroup'' only excludes one outgroup species from everyone else. %TODO

Figure~\ref{fig:4tabs}.d illustrates the traces of these four parameters against their logged states from MCMC chain. ``Human-Chimp'' and ``HomiCerco'' parameters are immediately reaching equilibrium from the starting point, but ``treeModel.rootHeight'' and ``ingroup'' clearly show a burn-in period. %TODO ?

\subsection*{Demographic reconstruction}

The second analysis uses an 1698bp alignment of 21 sequences from the Influenza A virus (?). %data citation
The piecewise-constant Bayesian skyline is selected for the tree prior, where the number of groups is reduced to 5 to avoid over-parameterize the model.
